---
format:
  html:
    embed-resources: true
    fig-height: 5
    fig-width: 8
    fig-dpi: 300
    toc: true
    toc-expand: 3
    toc-depth: 4
    toc-location: left
execute:
  keep-md: false
  cache: false
date: "`r Sys.Date()`"
editor: source
editor_options: 
  chunk_output_type: console
---


```{r library}
#| echo: false
#| message: false
library(dplyr)
library(glue)
library(readr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(forcats)
library(reactable)
library(gt)
library(stringr)
source('R/theme.R')
theme_set(theme_custom())
.data_global <- readr::read_csv('https://raw.githubusercontent.com/fbranda/measles/refs/heads/main/Measles_Global.csv')
.data_europe <- readr::read_csv('https://raw.githubusercontent.com/fbranda/measles/refs/heads/main/Measles_Europe.csv')

countries_of_interest <- c(
  'Canada',
  'United States of America', 'United Kingdom of Great Britain and Northern Ireland', 'Germany')
reactable <- function(...) {
  reactable::reactable(
    pagination = FALSE,
    searchable = TRUE,
    ...
  )
}
```


Example Topics

What is the growth rate of measles cases in the US vs Europe?

Are there particular counties or states at risk due to low vaccination rates?

How are vaccination rates changing over time?

What is your projection for the number of measles cases (and complications) in 2026?

# Europe vs global

growth rate

```{r}
.data_global |> glimpse()
.data_global |> count(Country)
.data_global_tidied <- .data_global |> 
  pivot_longer(
    any_of(month.name),
    names_to = 'month',
    values_to = 'count'
  )  |> 
  mutate(country_region = glue("{Country} - {Region}"))

.data_global_tidied_summarized <- .data_global_tidied |> 
  
  summarize(.by = c(Year, Country), count = sum(count, na.rm = TRUE)) 
.data_global_tidied_summarized |> 
  filter(Country %in% c('Canada', 'United States of America', 'Europe')) |> 
  ggplot(aes(x = Year, y = count, color = Country)) +
  geom_line() #+
#  guides(color = 'none')


.data_europe |> 
  glimpse()

.data_europe |>  count(HealthTopic, Population, Indicator)
.data_europe |>
  filter(Indicator == 'Reported confirmed cases') |> 
  summarize(.by = Time, count = sum(NumValue, na.rm = TRUE)) |> 
  mutate(Country = 'Europe') |>
  rename(Year = Time) |> 
  full_join(
    .data_global_tidied_summarized |> 
      filter(Country %in% c('Canada', 'United States of America', 'Europe'))) |> 
  ggplot(aes(x = Year, y = count, color = Country)) +
  geom_line() #+
#  guides(color = 'none')
```

```{r}

europe_countries <- .data_europe |>  pull(RegionName) |> unique()
setdiff(europe_countries, .data_global_tidied_summarized$Country)

.data_global_tidied_summarized |> 
  filter(
    Country %in% europe_countries |
      str_detect(Country, 'United Kingdom') |
      str_detect(Country, 'Netherlands') |
      Country %in% c('Canada', 'United States of America', 'Mexico')
  ) |>  
  mutate(region = case_when(
    
    
    Country %in% europe_countries |
      str_detect(Country, 'United Kingdom') |
      str_detect(Country, 'Netherlands')  ~ 'Europe',
    .default = Country
  )) |> 
  
  summarize(.by = c(Year, region), count = sum(count, na.rm = TRUE)) |> 
  ggplot(aes(x = Year, y = count, color = region)) +
  geom_line() #+

```


It seems that there is much more measles in Europe compared to USA. It is difficult
to interprety without adjusting for population numbers though.

```{r}

.data_global_tidied_summarized |> 
  filter(
    Country %in% europe_countries |
      str_detect(Country, 'United Kingdom') |
      str_detect(Country, 'Netherlands') |
      Country %in% c('Canada', 'United States of America', 'Mexico')
  ) |>  
  mutate(region = case_when(
    
    
    Country %in% europe_countries |
      str_detect(Country, 'United Kingdom') |
      str_detect(Country, 'Netherlands')  ~ 'Europe',
    .default = Country
  )) |> 
  
  ggplot(aes(x = Country, y = count, fill = as.factor(Year))) +
  geom_bar(stat = 'identity', position = position_dodge()) 


.data_global_tidied |> 
  mutate(date = glue("{Year}-{month}-01") |>  lubridate::ymd()) |> 
  filter(
    Country %in% c(
      'Canada',
      'United States of America', 'United Kingdom of Great Britain and Northern Ireland', 'Germany')
  ) |> 
  ggplot(aes(x = date, y = count, color = Country)) +
  geom_line() +
  scale_x_date(breaks = scales::pretty_breaks(n = 10))

```


# vaccination rate

1st or 2nd dose?

```{r}
.data_vac_rate <- readr::read_csv('https://raw.githubusercontent.com/fbranda/measles/refs/heads/main/Measles_vaccination_coverage_Global.csv')
.data_vac_rate |>  glimpse()
.data_vac_rate |> count(Antigen)
.data_vac_rate |> 
  filter(Country %in% countries_of_interest) |> 
  filter(Antigen == 'MCV1') |> 
  ggplot(aes(x = Year, y = Coverage, color = Country)) +
  geom_line()
```

missing usa data

# USA-specific

```{r}
data_usa_coverage <- readr::read_csv('https://raw.githubusercontent.com/fbranda/measles/refs/heads/main/USA/data/all/measles-USA-by-mmr-coverage.csv')
data_usa_all_by_year <- readr::read_csv('https://github.com/fbranda/measles/blob/main/USA/data/all/measles-USA-by-year.csv')

data_usa_by_state <- readr::read_csv('https://raw.githubusercontent.com/fbranda/measles/refs/heads/main/USA/data/2025/measles-USA-by-state-timeline.csv')

data_usa_coverage |> count(survey_type)

# convert
data_usa_coverage <- data_usa_coverage |> 
  mutate(coverage = str_remove(estimate_pct,'%') |> as.numeric(),
         coverage = coverage/100) |> 
  mutate(.by = geography,
         diff = max(coverage, na.rm = TRUE) - min(coverage, na.rm = TRUE))
data_usa_coverage |>
  distinct(geography, diff) |> 
  reactable()
data_usa_coverage |> 
  reactable()

# filter to top range
data_usa_coverage |> 
  ggplot(aes(x = school_year, y = coverage)) + 
  geom_line(aes(group = geography), color = 'grey') +
  geom_line(data = data_usa_coverage |> filter(diff > 0.1),
            aes(group = geography, color = geography), linewidth = 1.5) +
  scale_y_continuous(labels = scales::percent) +
  paletteer::scale_color_paletteer_d('ggsci::default_igv')

# to those states that have case counts in 2025
p <- data_usa_coverage |> 
  ggplot(aes(x = school_year, y = coverage)) + 
  geom_line(aes(group = geography), color = 'grey') +
  geom_line(
    data = data_usa_coverage |> 
      filter(geography %in% data_usa_by_state$state_name),
    aes(group = geography, color = geography), linewidth = 1.5) +
  scale_y_continuous(labels = scales::percent) +
  paletteer::scale_color_paletteer_d('ggsci::default_igv')  ; 
plotly::ggplotly(p)
```

## usa MMR rate over year
mean over states
```{r}
get_stats <- function(x) {
  tibble(mean = mean(x, na.rm = TRUE),
         median = median(x, na.rm = TRUE))
}
data_usa_coverage |> 
  reframe(
    .by = c(school_year), 
    get_stats(coverage),
    n = sum(!is.na(coverage))
  ) |> 
  pivot_longer(c(mean, median), names_to = 'stat') |> 
  ggplot(aes(x = school_year, y = value, group = stat, color = stat)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent, expand = expansion(0.1, 0.01))  

```

## 2025 cases

```{r}
data_usa_by_state |> 
  ggplot(aes(x = report_date, y = cases_count, color = state_name)) +
  geom_line() +
  geom_point(aes()) +
  paletteer::scale_color_paletteer_d('ggsci::default_igv', guide = 'none') +
  facet_wrap(vars(state_name))+
  theme(panel.border = element_rect(fill = NA))
```

